---
title: "Correspondence Analysis of OWB Import Data — PG32 Parameter"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 7
)

library(here)
library(almanac)
library(CA3variants)
library(openxlsx)
library(knitr)
```

## Overview

This report presents a three-way Non-Symmetric Correspondence Analysis (NSCA3)
examining the association between:

- **Trap result** (row / term I): whether an OWB trap collection occurred within
  a given monitoring window
- **Country of origin** (column / term J): country associated with the imported
  shipment
- **Commodity** (tube / term K): commodity description from CBP/HTS data

Data source: `01-imports_PG32_origin_2026.02.04.csv`

---

## Load Processed Data

```{r load-data}
pg32Data <- readRDS(here::here("data", "processed", "pg32Data.rds"))

formatData.pg32.AllTraps <- readRDS(
  here::here("data", "processed", "formatData_pg32_AllTraps.rds")
)

kable(head(formatData.pg32.AllTraps, 10), caption = "First 10 rows of formatted data")
```

---

## Build the Contingency Table

```{r contingency-table}
contingency.pg32.AllTraps <- table(formatData.pg32.AllTraps[, c(2, 3, 4)])

cat("Contingency table dimensions:\n")
cat("  Trap result (I) x Country (J) x Commodity (K) =",
    paste(dim(contingency.pg32.AllTraps), collapse = " x "), "\n")
```

---

## Tune the Number of Dimensions

The `tunelocal` function helps identify the optimal number of dimensions for
the NSCA3 model. The response variable is the row (trap result).

```{r tune-dimensions}
tune.pg32.AllTraps <- tunelocal(contingency.pg32.AllTraps,
                                ca3type = 'NSCA3',
                                resp    = 'row')
plot(tune.pg32.AllTraps)
```

---

## Fit the NSCA3 Model

Based on the tuning plot, dimensions `c(2, 4, 4)` are selected:

- **2 dimensions** for the trap-result (row) component
- **4 dimensions** for country (column) and commodity (tube) components

Increasing the column/tube dimensions to 5 pushes explained inertia above the
70 % threshold recommended by Beh & Lombardo.

```{r fit-model}
ca3.pg32.AllTraps <- CA3variants(contingency.pg32.AllTraps,
                                  ca3type = 'NSCA3',
                                  resp    = 'row',
                                  dims    = c(2, 4, 4))
ca3.pg32.AllTraps
```

---

## Biplots

### Row component (Trap result)

```{r biplot-row}
plot(ca3.pg32.AllTraps, biptype = "row", addlines = FALSE, scaleplot = 15)
```

### Column component (Country)

```{r biplot-col}
plot(ca3.pg32.AllTraps, biptype = "col", addlines = FALSE, scaleplot = 15)
```

### Tube component (Commodity)

```{r biplot-tube}
plot(ca3.pg32.AllTraps, biptype = "tube", addlines = FALSE, scaleplot = 15)
```

---

## Inner-Product Matrix

The inner-product matrix indicates the strength of the association between
trap-result categories, countries, and commodities.

```{r inner-products}
kable(t(ca3.pg32.AllTraps$iproductijk),
      caption = "Inner-product matrix (transposed)")
```

---

## Save Output

```{r save-output}
myToday           <- as.character(Sys.Date())
outPrefixFileName <- 'correspondence_commodity_country_2.4.4dims'
outfile           <- paste0(outPrefixFileName, '_', myToday, '.xlsx')
outDirFile        <- here::here("output", "data", outfile)

wb <- createWorkbook(outPrefixFileName)
addWorksheet(wb, sheetName = myToday)
writeData(wb, myToday,
          x        = t(ca3.pg32.AllTraps$iproductijk),
          rowNames = TRUE)
saveWorkbook(wb, file = outDirFile, overwrite = TRUE)

cat("Output saved to:", outDirFile, "\n")
```

---

## Session Info

```{r session-info}
sessionInfo()
```
